---
title: "Final Project: Understanding what Factors are Associated with Higher Life Expectancy"
author: Mathew Joseph Jacob
output:
  pdf_document: default
  html_notebook: default
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}

#### SETUP #####################################################################

# Imports necessary libraries
library("gapminder")
library("ggplot2")
library("dplyr")
library("stargazer")
library(knitr)

# Imports necessary data sources and formats chunks
data("gapminder")
gapminder <- data.frame(gapminder)
worldbankdata <- read.csv("worldbankdata.csv", as.is = T)
opts_chunk$set(tidy.opts=list(width.cutoff=60))   



#### DATA CLEANING #############################################################

# Fixes country names in gapminder data to match the World Bank Country Names
corrections <- read.csv("corrections.csv", as.is = T)
countries = c()
for (country in gapminder$country){
  if(country %in% corrections$wrong){
    correct <- corrections[corrections$wrong == country,]$right
    countries <- c(countries, correct)
  }else{
    countries <- c(countries, country)
  }
}
gapminder$country <- countries

# Gets list of all country names and regions
countryinfo <- gapminder[!duplicated(gapminder$country),]
countryinfo <- countryinfo[ ,-c(3:6)]

# Creates GDP dataframe from Gapminder Data
gdpPC <- gapminder
gdpPC[, c(2,4,5)] <- NULL
gdpPC <- reshape(gdpPC, direction = "wide", idvar = "country", timevar = "year")
gdpPC <- merge(countryinfo, gdpPC, by = "country", all.x = T, all.y = T)
rownames(gdpPC) <- NULL

# Creates Life Expectancy dataframe from Gapminder Data 
lfExp <- gapminder
lfExp[, c(2,5,6)] <- NULL
lfExp <- reshape(lfExp, direction = "wide", idvar = "country", timevar = "year")
lfExp <- merge(countryinfo, lfExp, by = "country", all.x = T, all.y = T)
rownames(lfExp) <- NULL

# Creates PM 2.5 Exposure (microgram/m^3) dataframe from Worldbank Data
pollution <- worldbankdata[worldbankdata$Series.Code == "EN.ATM.PM25.MC.M3", ]
pollution[, c(2,3,4)] <- NULL
pollution <- data.frame(pollution[, 1], apply(pollution[,-1], 2, as.numeric))
colnames(pollution) <- c("country", as.character(1960:2019))

# Creates Sanitation Access dataframe from Worldbank Data
sanitation <- worldbankdata[worldbankdata$Series.Code == "SH.STA.BASS.ZS", ]
sanitation[, c(2,3,4)] <- NULL
sanitation <- data.frame(sanitation[, 1], apply(sanitation[,-1], 2, as.numeric))
colnames(sanitation) <- c("country", as.character(1960:2019))

# Creates Drinking Water Access dataframe from Worldbank Data
water <- worldbankdata[worldbankdata$Series.Code == "SH.H2O.BASW.ZS", ]
water[, c(2,3,4)] <- NULL
water <- data.frame(water[, 1], apply(water[,-1], 2, as.numeric))
colnames(water) <- c("country", as.character(1960:2019))



#### DATA MERGING ##############################################################

# Gets 2007 Sanitation Data
sanitation2007 <- data.frame(sanitation$country, as.numeric(sanitation$`2007`))
colnames(sanitation2007) <- c("country","sanitation2007")

# Gets 2010 Pollution Data
pollution2010 <- data.frame(pollution$country, pollution$`2010`)
colnames(pollution2010) <- c("country","pollution2010")

# Gets 2007 Water Data
water2007 <- data.frame(pollution$country, as.numeric(water$`2007`))
colnames(water2007) <- c("country","water2007")

# Combines all data
merged <- data.frame(lfExp$country, lfExp$continent, 
                     lfExp$lifeExp.2007, gdpPC$gdpPercap.2007)
colnames(merged) <- c("country","continent","lifeExp2007","gdpPerCap2007")
merged <- merge(merged, sanitation2007, by = "country")
merged <- merge(merged, pollution2010, by = "country")
merged <- merge(merged, water2007, by = "country")
merged$gdpPerCap2007 <- log(merged$gdpPerCap2007)

# Gets China Gapminder Data
Chinagapminder <- gapminder[gapminder$country == "China", ]

# Gets China Pollution Data
Chinapollution <- pollution[pollution$country == "China", -c(2:41)]
Chinapollution <- reshape(Chinapollution, direction = "long", 
                       varying = list(names(Chinapollution)[2:21]),
                       v.names = "value", idvar = c("country"),
                       timevar = "year", times = 2000:2019)

# Gets China Sanitation Data
Chinasanitation <- sanitation[sanitation$country == "China", -c(2:41)]
Chinasanitation <- reshape(Chinasanitation, direction = "long", 
                       varying = list(names(Chinasanitation)[2:21]), 
                       v.names = "value", idvar = c("country"),
                       timevar = "year", times = 2000:2019)

# Gets China Water Data
Chinawater <- water[water$country == "China", -c(2:41)]
Chinawater <- reshape(Chinawater, direction = "long", 
                        varying = list(names(Chinawater)[2:21]),
                        v.names = "value", idvar = c("country"),
                        timevar = "year", times = 2000:2019)
```

<!--  
This is a template for creating your final project report. It lays out the sections that should be in your write-up and describes a little about these sections. There is some flexibility to deviate from this structure, for example, interweaving more visualizations and analyses could work well. 

Your report should be between 5-8 pages long and should contain:

    1) Introduction: 
      a) what is question you are addressing 
      b) why is important 
      c) where you got the data
      d) description of any other analyses that been done on the data
      
    2) Visualizations of the data: one or more plots
    
    3) Analyses: models, hypothesis tests, confidence intervals and other inferential statistics that give 
       insight into your question
    
    4) Conclusions: What you found, future directions, etc.
    
    5) Reflection (do be completed on Canvas)
       a) What went well 
       b) What did you struggle with
       c) What analyses did you do that you are not including, etc. 

Please make your report look good by paying attention to detail, using additional R Markdown features etc. 

If there is additional code or data you would like to include with your report, please create a github page and add a link to it in your report. Additionally, you can append the full code for your analysis in an appendix section at the end of the document, and then include only the most important pieces of code in the body of the report. For example, you can exclude details of data cleaning from the body of the report, but include anything of central to your analyses, and also any information about particular choices you made that might affect the results and why you made those choices (e.g., explain if data was excluded and why, etc.). 
--> 



## Introduction 

<!--  
Write ~1-3 paragraphs describing:

1) What is problem you are addressing and why the problem you are addressing is interesting

2) Where you got the data from, including a link to the website where you got the data if applicable 

3) What other analyses have already been done with the data and possibly links to other analyses, also mention if you are using the data in another class or for another research project. 
--> 

Many countries are interested in improving the life expectancy of citizens. However, it is important to know which factors are associated with life expectancy. Using data from the Gapminder data set and data from the World Bank, this analysis will look to find if improvements in water access, decreasing exposure to pollutants (using PM 2.5 as a proxy), or improving access to proper sanitation is associated with a larger improvement in life expectancy. The gapminder package data was used to provide data on a countries Life Expectancy and GDP per Capita (which was used experimented with as a control variable in the regression). Water access, polution exposure, and sanitation access data were obtained from the World Bank. By going to this *[link](https://databank.worldbank.org/source/world-development-indicators#)* and selecting these indicators, the necessary data was downloaded as one single csv file which was renamed as worldbankdata.csv. The data used in this analysis has not been used for other classes or research projects.

* *[Find the water access data here.](https://data.worldbank.org/indicator/SH.H2O.SMDW.ZS)*

* *[Find the sanitation access data here.](https://data.worldbank.org/indicator/SH.STA.SMSS.ZS)*

* *[Find the PM 2.5 emissions access data here .](https://data.worldbank.org/indicator/EN.ATM.PM25.MC.M3)*

A *[study](Socio-economic correlates of life expectancy at birth--the case of developing countries.)* on a similar topic was done in 1990 that looked into other factors that corressponded with Life Expectancy. However, aside from water access, this study focused on differeent factors such as energy consumption/capita, literacy rates, and family planning. I worked on this project alone.



## Data wrangling

<!--  
Very briefly discuss how you got the data into shape for your analyses. You can include some code here, although extensive data cleaning code should be put on github or in an appendix at the end of the document.
--> 

### Gapminder Data Cleaning
First, several of the country names in the gapminder dataset were changed to match the names used by the World Bank to prevent these countries from being removed when the data sets were merged. The GDP per Capita (control variable) and Life Expectancy (dependent variable) data were put into their own data frames and pivoted. Plotting Life Expectancy against GDP per Capita resulted in a log pattern and so a log transformation was applied to the GDP per Capita data.

### World Bank Data Cleaning
For each variable (water access, sanitation access, and polution exposure), the same steps were applied to clean the data. First, the data relevant to one particular variable was filtered from the worldbankdata.csv. Unnecessary columns were dropped, the numerical values were converted from Strings to numbers and absent values were set to NA. Finally, the columns were renamed to make it easier to access the data for a particular year. In order to make creating the line plots for these three variables easier, the variable dataframes were pivoted.

### Merged Data
The GDP per Capita, Life Expectancy, water access, and sanitation access data from 2007 and the polution exposure data from 2010 (the closest year available) were merged into one data set for the regression analysis.



## Visualizations: Interest variables over time and understanding GDP per Capita

<!--  
Create one or more plots of your data. Describe the plot(s), what they shows, and why they are of interest to your analysis. Include the code to create these plots in the R chunk below. You can also discuss the plots after the code too. Finally, be sure to make plots as clear as possible (clear axis labeling, legends and captions) so that it is easy for the reader to quickly understand the central information being conveyed. 
--> 

```{r message=FALSE, warning=FALSE, tidy=TRUE, fig.width=3.70, fig.height=3.70, fig.align='center'}

# Plots China GDP Over Time
ggplot(Chinagapminder) + geom_line(aes(x = year, y = gdpPercap)) + 
  ggtitle("China GDP per Capita Over Time") + xlab("Year") + ylab("GDP Per Capita")

# Plots China Life Expectancy Over Time
ggplot(Chinagapminder) + geom_line(aes(x = year, y = lifeExp)) + 
  ggtitle("Life Expectancy Over Time") + xlab("Year") + ylab("Life Expectancy")

# Plots China PM 2.5 Exposure Over Time
ggplot(Chinapollution) + geom_line(aes(x = year, y = value)) + 
  ggtitle("China PM 2.5 Exposure Over Time") + xlab("Year") + ylab("PM2.5 Exposure (micrograms per m^3")

# Plots China Sanitation Access Over Time
ggplot(Chinasanitation) + geom_line(aes(x = year, y = value)) + 
  ggtitle("China Sanitation Access Over Time") + xlab("Year") + ylab("% of Population")

# Plots China Drinking Water Access Over Time
ggplot(Chinawater) + geom_line(aes(x = year, y = value)) +
  ggtitle("China Drinking Water Access Over Time") + 
  xlab("Year") + ylab("% of Population")
```

```{r message=FALSE, warning=FALSE, tidy=TRUE, fig.width=5, fig.height=4.5, fig.align='center'}

# Plot GDP per Capita vs LFE
plot(gdpPC$gdpPercap.2007, lfExp$lifeExp.2007, 
     main = "GDP Per Capita vs Life Expectancy", 
     xlab = "GDP Per Capita", ylab = "Life Expectancy")
```

### Visualization Discussion
To get a better idea of the variables, the following plots were created using China as an example country. In the first two plots the data available went back to 1952. It shows that aside from a drop in 1960, China has been improving in these two categories over time. For the next three plots of Pollution Exposure, Sanitation Access, and Water Access, data was only available from 2000 onwards and the plots were restricted to this time frame to make it easier to understand. China saw improvements in Pollution Exposure (decreasing), and Sanitation and Water Access (increasing)

The final plot of GDP per Capita against LFE was used to show the logarithmic pattern in order to justify a log transformation of this variable.


## Analyses: Regression Models

<!--  
Build linear models, run hypothesis tests, create confidence intervals and/or run simulations to answer questions that are of interest. 
--> 

### First Set of Regression Models
```{r results='asis', fig.align='center'}

model1 <- lm(lifeExp2007 ~ pollution2010, data = merged)

model2 <- lm(lifeExp2007 ~ water2007, data = merged)

model3 <- lm(lifeExp2007 ~ sanitation2007, data = merged)

stargazer(model1, model2, model3, header = F, title = "Table 1", omit.stat=c("f"), float = FALSE)
```

### Second Set of Regression Models
```{r results='asis', fig.align='center'}

model4 <- lm(lifeExp2007 ~ pollution2010 + gdpPerCap2007 , data = merged)
model4$AIC <- AIC(model4)
model4$BIC <- BIC(model4)  

model5 <- lm(lifeExp2007 ~ water2007 + gdpPerCap2007 , data = merged)
model5$AIC <- AIC(model5)
model5$BIC <- BIC(model5)

model6 <- lm(lifeExp2007 ~ sanitation2007 + gdpPerCap2007 , data = merged)
model6$AIC <- AIC(model6)
model6$BIC <- BIC(model6)

model7 <- lm(lifeExp2007 ~ water2007 + sanitation2007 + gdpPerCap2007 , data = merged)
model7$AIC <- AIC(model7)
model7$BIC <- BIC(model7)

stargazer(model4, model5, model6, model7, header = F, title = "Table 2", omit.stat=c("f"), float = FALSE)
```


## Conclusion 

<!--  
1-2 paragraphs summarizing what you found, how the findings address your question of interest, and possible future directions. Please make sure describe your conclusions in an intuitive way, and make sure that your argument is strong and backed by solid evidence from your data. 
-->

Based on the first set of regressions without controling for GDP per Capita (Table 1), as shown by the 3 stars (a p-value below 0.01) next to the coefficients of each model. The coefficient signs also make sense as increasing the amount of pollution leads to a decrease in life expectancy for the first model. On the other hand increasing the percent of the population that has access to sanitation services and clean water leads to an increase in life expectancy for the second and third model. However, while all coefficients are significant, they do not explain a comparable amount. Comparing the $R^2$ is a proper indication of this since we are not concerned about overfitting when we are only using one variable values (making adjusted $R^2$, AIC, and BIC less necessary to utilize). The $R^2$ for model 1 shows that only 0.125 percent of the variability in life expectancy is explained by pollution. Water and sanitation access on the other hand explain 0.736 percent and 0.754 percent respecctively of the variability in life expectancy.

However, because countries can be very different, adding a control variable such as GDP per Capita can help provide more information. When this variable is added to the models (Table 2), it can be seen that pollution is no longer significant. However, water and sanitation access are still significant at the 0.01 level and the $R^2$ increases as well for both models. Finally, creating a model with water and sanitation access and GDP per Capita resulted in a model where all three coefficients were signficant at least at the 0.05 level. The $R^2$ increased as well compared to the models with only sanitation or water access and GDP per Capita. However, since multiple variables are being used measures such as adjusted $R^2$, AIC, and BIC are better for making sure this addition is meaningful. All three scores indicate this final model is the best at explaining the variability of life expectancy.

This analysis shows that increasing water and sanitation access are associated with a higher life expectancy and that exposure to pollution (using PM 2.5 as a proxy) is not as signficant of a predictor. An important distinction is that this does not necessarily indicate causality. Experiments or using more complicated analyses (such as setting up a difference in differences study) are possible future areas to better understand whether water and sanitation access are a direct causal factor of a countries' life expectancy.


## Reflection

<!--  
Reflection  

Write one paragraph describing what went well with this project and what was more difficult. Also describe any additional things you tried that you did not end up including in this write-up, and approximately how much time you spend working the project. 

Finally, please go to Canvas and answer a few questions related to how this project went: https://yale.instructure.com/courses/51220/quizzes/20279
-->

I thought this project was a very helpful exercise in applying the concepts we learned in class. I felt that creating and comparing the regressions and visualizing the data for China as an example over time went well. I felt the most difficult part was cleaning the data because there were countries with different names from the different sources. Reshaping the data to make it easier for merging or creating a time series was also tricky. I also tried experimenting with other factors such as environmental health but did not include these in the final write-up because the data I found did not have as many of the countries that the gapminder package had and would have resulted in models comparing variables with a significantly different number of data points. I spent 5.5 hours on the analysis and 2 hours writing up the report.

## Appendix

<!--  
You can include a complete listing of your code here if you could not fit it into the body of the document. Make sure your code is well commmented and easy to create - i.e., use meaningful object names, separate your code into sections, describe what each section is doing, etc. 
-->

```{r eval=FALSE}

# Complete code for the entire analysis.

#### SETUP #####################################################################

# Imports necessary libraries
library("gapminder")
library("ggplot2")
library("dplyr")

# Imports necessary data sources
data("gapminder")
gapminder <- data.frame(gapminder)
worldbankdata <- read.csv("worldbankdata.csv", as.is = T)



#### DATA CLEANING #############################################################

# Fixes country names in gapminder data to match the World Bank Country Names
corrections <- read.csv("corrections.csv", as.is = T)
countries = c()
for (country in gapminder$country){
  if(country %in% corrections$wrong){
    correct <- corrections[corrections$wrong == country,]$right
    countries <- c(countries, correct)
  }else{
    countries <- c(countries, country)
  }
}
gapminder$country <- countries

# Gets list of all country names and regions
countryinfo <- gapminder[!duplicated(gapminder$country),]
countryinfo <- countryinfo[ ,-c(3:6)]

# Creates GDP dataframe from Gapminder Data
gdpPC <- gapminder
gdpPC[, c(2,4,5)] <- NULL
gdpPC <- reshape(gdpPC, direction = "wide", idvar = "country", timevar = "year")
gdpPC <- merge(countryinfo, gdpPC, by = "country", all.x = T, all.y = T)
rownames(gdpPC) <- NULL

# Creates Life Expectancy dataframe from Gapminder Data 
lfExp <- gapminder
lfExp[, c(2,5,6)] <- NULL
lfExp <- reshape(lfExp, direction = "wide", idvar = "country", timevar = "year")
lfExp <- merge(countryinfo, lfExp, by = "country", all.x = T, all.y = T)
rownames(lfExp) <- NULL

# Creates PM 2.5 Exposure (microgram/m^3) dataframe from Worldbank Data
pollution <- worldbankdata[worldbankdata$Series.Code == "EN.ATM.PM25.MC.M3", ]
pollution[, c(2,3,4)] <- NULL
pollution <- data.frame(pollution[, 1], apply(pollution[,-1], 2, as.numeric))
colnames(pollution) <- c("country", as.character(1960:2019))

# Creates Sanitation Access dataframe from Worldbank Data
sanitation <- worldbankdata[worldbankdata$Series.Code == "SH.STA.BASS.ZS", ]
sanitation[, c(2,3,4)] <- NULL
sanitation <- data.frame(sanitation[, 1], apply(sanitation[,-1], 2, as.numeric))
colnames(sanitation) <- c("country", as.character(1960:2019))

# Creates Drinking Water Access dataframe from Worldbank Data
water <- worldbankdata[worldbankdata$Series.Code == "SH.H2O.BASW.ZS", ]
water[, c(2,3,4)] <- NULL
water <- data.frame(water[, 1], apply(water[,-1], 2, as.numeric))
colnames(water) <- c("country", as.character(1960:2019))



#### DATA MERGING ##############################################################

# Gets 2007 Sanitation Data
sanitation2007 <- data.frame(sanitation$country, as.numeric(sanitation$`2007`))
colnames(sanitation2007) <- c("country","sanitation2007")

# Gets 2010 Pollution Data (nearest year available)
pollution2010 <- data.frame(pollution$country, pollution$`2010`)
colnames(pollution2010) <- c("country","pollution2010")

# Gets 2007 Water Data
water2007 <- data.frame(pollution$country, as.numeric(water$`2007`))
colnames(water2007) <- c("country","water2007")

# Combines all data
merged <- data.frame(lfExp$country, lfExp$continent, 
                     lfExp$lifeExp.2007, gdpPC$gdpPercap.2007)
colnames(merged) <- c("country","continent","lifeExp2007","gdpPerCap2007")
merged <- merge(merged, sanitation2007, by = "country")
merged <- merge(merged, pollution2010, by = "country")
merged <- merge(merged, water2007, by = "country")



#### VISUALIZATION #############################################################

# Gets China Gapminder Data
Chinagapminder <- gapminder[gapminder$country == "China", ]

# Gets China Pollution Data
Chinapollution <- pollution[pollution$country == "China", -c(2:41)]
Chinapollution <- reshape(Chinapollution, direction = "long", 
                       varying = list(names(Chinapollution)[2:21]),
                       v.names = "value", idvar = c("country"),
                       timevar = "year", times = 2000:2019)

# Gets China Sanitation Data
Chinasanitation <- sanitation[sanitation$country == "China", -c(2:41)]
Chinasanitation <- reshape(Chinasanitation, direction = "long", 
                       varying = list(names(Chinasanitation)[2:21]), 
                       v.names = "value", idvar = c("country"),
                       timevar = "year", times = 2000:2019)

# Gets China Water Data
Chinawater <- water[water$country == "China", -c(2:41)]
Chinawater <- reshape(Chinawater, direction = "long", 
                        varying = list(names(Chinawater)[2:21]),
                        v.names = "value", idvar = c("country"),
                        timevar = "year", times = 2000:2019)

# Plots China GDP Over Time
ggplot(Chinagapminder) + geom_line(aes(x = year, y = gdpPercap)) + 
  ggtitle("China GDP Over Time") + 
  xlab("Year") + ylab("GDP Per Capita")

# Plots China Life Expectancy Over Time
ggplot(Chinagapminder) + geom_line(aes(x = year, y = lifeExp)) + 
  ggtitle("Life Expectancy Over Time") + 
  xlab("Year") + ylab("Life Expectancy")

# Plots China PM 2.5 Exposure Over Time
ggplot(Chinapollution) + geom_line(aes(x = year, y = value)) + 
  ggtitle("China PM 2.5 Exposure Over Time") + 
  xlab("Year") + ylab("PM 2.5 Exposure (micrograms per cubic meter")

# Plots China Sanitation Access Over Time
ggplot(Chinasanitation) + geom_line(aes(x = year, y = value)) + 
  ggtitle("China Sanitation Access Over Time") + 
  xlab("Year") + ylab("% of Population")

# Plots China Drinking Water Access Over Time
ggplot(Chinawater) + geom_line(aes(x = year, y = value)) + 
  ggtitle("China Drinking Water Access Over Time") + 
  xlab("Year") + ylab("% of Population")

# Plot GDP vs LFE and log transform GDP
plot(gdpPC$gdpPercap.2007, lfExp$lifeExp.2007, 
     main = "GDP Per Capita vs Life Expectancy", 
     xlab = "GDP Per Capita", ylab = "Life Expectancy")
merged$gdpPerCap2007 <- log(merged$gdpPerCap2007)


#### REGRESSION ANALYSIS #######################################################

model1 <- lm(lifeExp2007 ~ pollution2010, data = merged)

model2 <- lm(lifeExp2007 ~ water2007, data = merged)

model3 <- lm(lifeExp2007 ~ sanitation2007, data = merged)

model4 <- lm(lifeExp2007 ~ pollution2010 + gdpPerCap2007 , data = merged)
model4$AIC <- AIC(model4)
model4$BIC <- BIC(model4)  

model5 <- lm(lifeExp2007 ~ water2007 + gdpPerCap2007 , data = merged)
model5$AIC <- AIC(model5)
model5$BIC <- BIC(model5)

model6 <- lm(lifeExp2007 ~ sanitation2007 + gdpPerCap2007 , data = merged)
model6$AIC <- AIC(model6)
model6$BIC <- BIC(model6)

model7 <- lm(lifeExp2007 ~ water2007 + sanitation2007 + gdpPerCap2007 , data = merged)
model7$AIC <- AIC(model7)
model7$BIC <- BIC(model7)

```

